---
title: "Predictive Modelling for Loans"
output: html_notebook
---
```{r}
library('readxl')
library('tidyverse')
library('janitor')
library('GGally')
```

```{r}
library(readr)
lending_club_loans <- read_csv("C:/Users/44792/OneDrive/Desktop/Codeclan 2022/final_project_loan_default_predictor/data/lending_club_loans.csv")
View(lending_club_loans)
```
## Background
#Lending Club was established in 2007 and there are now over 4 million members. It is a #peer-to-peer lender and the first in the USA to register its offering as securities #with SEC (Securities and Exchange Commission).
#Lending Club, or other P2 P lenders, intersects technology and finance and with this #approach there is the possibility of progress a more effective business model and using #machine learning to develop a risk management model that can eliminate lending #bias(introduced by humans).  This data and subsequent model is a good example of how DDD #(Data Driven Decisions) can be used to ensure there is a more transparent approach to #accessing personal finance.
#Data set is from Kaggle and is collected from 2007-2018.

# Ethical consideration in the development of Machine Learning Models

#When gathering person data consumer data protection needs to be considered, respected and the legislation followed.
#Secondly, when considering the use of machine learning it is important to be transparent #about the use of machine learning during the decision-making process- as noted in annual #reports from LC marketing materials are clear that the company use over 150million cells #of data to support the use of artificial intelligence-driven credit decisions and #machine learning models are being used across the customer lifecycle to expand access to #credit
#Diversity and inclusion in the development of a model is also important as bias can be #introduced if the machine learning models are limited by the data that is gets to train #on.  Developing a team with a wide range of experience and from a diverse range of #backgrounds is important.

#I would recommend that further work be done on some of the new test models “What If” by Google so that the models can be challenged using a number of other inputs.


##Data Preparation and Cleaning
```{r}
view(lending_club_loans)
lending_club_loans <- clean_names(lending_club_loans)

```
```{r}
loans1 <- lending_club_loans %>%
  select(loan_amnt,term,int_rate,sub_grade,home_ownership,annual_inc,purpose, loan_status,dti)
```


#View the structure of the data
```{r}
glimpse(loans1)
```
```{r}
str(loans1)
```


```{r}
summary(loans1)
```


```{r}
loans2 <- loans1 %>%
  filter(!is.na(loan_amnt))
loans2

loans3 <- loans2 %>%
  filter(!is.na(annual_inc))

```


```{r}
summary(as.factor(loans2$term))
```
#Majority of the loans are for 36 months (74%) and 26% and for 60 months.

```{r}
summary(as.factor(loans2$sub_grade))
```
# The Lending Club internal credit rating ranges from A to G with 5 subcategories (a total of 35 credit ratings). A1 being the least likely to default and G5 the catergory with the highest risk profile


```{r}
summary(as.factor(loans2$dti))
```

```{r}
summary(as.factor(loans2$purpose))
```
#Debt consolidation (19,776 loans) 46% of loans followed by paying off credit card debt (5,477) 13% and then other (4425) 10% are the top three purposes of the loan request. It should be noted that this is self reported by the applicant and caution should be exercised when considering the validity of this informations.

#While debt consolidation could arise from various other sources, such as auto or home equity lines, loans from these sources are secured and, hence, considerably different than unsecured credit like LC

```{r}
summary(as.factor(loans2$loan_status))
```


```{r}
glimpse(loans3)
```
#Further cleaning of the data here at this point to drop the other loan status information.

```{r}
loansdropped <-loans3 %>%
  filter(!loan_status %in% c("Current","Default",
                    "Does not meet the credit policy. Status:Charged Off",
                    "Does not meet the credit policy. Status:Fully Paid ",
                    "In Grace Period",
                    "In Grace Period",
                    "Late (31-120 days) "))
loansdropped
```


## Explority analysis

##Before building a model i think it is important to consider the key characteristics of a loan that has been defaulted on so that we are in a stronger position to understand what are the strong predictors of aloan defaulting.


#Visual and explore the characteristics of the defaults loans only to uderstand more about the type of loan that is written off.




```{r}
default1 <-loansdropped %>%
  filter(loan_status == "Charged Off")
default1
```

```{r}
glimpse(default1)
```
```{r}
summary(default1)
```
#Average value of the defaulted loan is $12,134

```{r}
default1 %>%
  group_by(term, term_count = n())
```


```{r}
default_arranged <- default1 %>%
  select(loan_amnt, term)%>%
  arrange(desc(loan_amnt))
```

```{r}
summary(default_arranged)
```
#Data set for default loans is 5,653 and on average the default amount is $12,134 - across the business and the timeline of the data this could be a value of $68,593,502 so any reduction in the number of defaulting loans by 1% would be $68,000 additional revenue.
##Total of 42,531 individual loan accounts with 5,653 charged off (meaning that the lender has written off the debt and the account is closed) 
#(13% of the total loans made) - so a reduction in the defualts of 1% (is across this data the average losses were $68m - this )

#Total of 33,586 fully paid (79% of the total loans)
#From the data set there are some records that are unclear catergories

```{r}
ggplot(default1) +
  geom_bar(aes(x= term), fill = "light blue")+
labs( x= "Term of Loan",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data")
```
# Loans with the longer term of 60 months make up a higher percentage of the defaults loans - so we could conclude that the longer the  term the higher the exposure to changes in a persons financial status e.g employment. The loan term could be considered a factor in the probability of a customer defaulting.

```{r}
ggplot(default1) +
  geom_bar(aes(x= loan_amnt), fill = "light blue")+
  coord_flip()+
labs( x= "Loan Amount",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data")
```
#Here you can see four spikes in the data where there are more defaults on certain values of loans - $7,000, $18,000, $25,000 and then the max of $35,000 - the loan size could be considered a factor in the probabilty of a customer defaulting.



```{r}
ggplot(default1) +
  geom_bar(aes(x=home_ownership), fill = "light blue")+
   labs( x= "Houseing Status",
        y= "Number of Default Loans",
        title = "Lending Club Loans Data",
        subtitle = "Houseing Status")
```
#Default loans also have a higher percentage of renters - unsurprising and also might be considered a higher risk when lending as a credit check cannot be secured from the other lender. Furthermore those with a mortage may be in a position to get a loans that are secured against their property. Housing status could be considered a factor in the probability of a customer defaulting.


```{r}
ggplot(default1) +
  geom_bar(aes(x= term, fill = purpose))+
  labs( x= "Term of Default Loan",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data",
        subtitle = "Term & Purpose",
        fill = "Purpose of Loan")
```
#Debt consolidation, payment of credit card debts and "other" continue to be the top three reasons for a loan and default.  The purpose of a loan could be considered a factor in the chance of a customer defaulting.

#Limiting exposure to the loans being used for this purpose may be a consideration in the future.



```{r}
ggplot(default1) +
  geom_bar(aes(x= int_rate, fill = term))+
  labs( x= "Interest Rate",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data",
        subtitle = "Term & Purpose",
        fill = "Term")
```
XX rubbish plot - may chuck



  

```{r}
ggplot(default1) +
  geom_bar(aes(x= term, fill = annual_inc))+
  labs( x= "Term of Default Loan",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data",
        subtitle = "Term & Purpose",
        fill = "Purpose of Loan")
```




```{r}

```






#loan status needs to be turned into a numerical column 0 or 1 for purpose of building a predictive model.

#Our ultimate goal is the prediction of loan defaults from a given set of observations by selecting explanatory (independent) variables (also called feature in machine learning parlance) that result in an acceptable model performance as quantified by a pre-defined measure. This goal will also impact our exploratory data analysis. 



```{r}
loansmodelling <-loansdropped %>%
  mutate(loan_status = recode(loan_status,"Fully Paid" = "1", "Charged Off" = "0")) 
loansmodelling
```



```{r}
glimpse(loansmodelling)
```



```{r}
head(loansmodelling, 15)
```














#Machine Model



#Now let's have a look at the relationships with loan status. I will split the variables into three sets, as the output of `ggpairs()` will be slow and cramped if we include all variables at once. Note that a disadvantage of this approach is that it will be difficult to visually assess relationships between some predictor pairs if they fall into different split sets.


```{r}
loans_split1 <- loans4 %>%
  select(loan_amnt, term,home_ownership, loan_status)
```


```{r}
loans_split2 <- loans4 %>%
  select(annual_inc, purpose,loan_status)
```

#Now develop the connection with ggpairs

```{r}
library(GGally)
loans_split1 %>%
  ggpairs()
```

```{r}
library(GGally)
loans_split2 %>%
  ggpairs()
```









