---
title: "Lending Club - Identifying Factors that Influence Customers Defaulting on
  Loans"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---
```{r}
library('readxl')
library('tidyverse')
library('janitor')
library('GGally')
library('fastDummies')
library('ggfortify')
library('mosaic')
library('mosaicData')
library('modelr')
library('readxl')
library('broom')
library("wesanderson")
library("RColorBrewer")
```


```{r}
library(readr)
lending_club_loans <- read_csv("C:/Users/44792/OneDrive/Desktop/Codeclan 2022/final_project_loan_default_predictor/data/lending_club_loans.csv")
View(lending_club_loans)
```
#   Background

####  Lending Club (LC) was established in 2007 and there are now over 4 million members. It is a peer-to-peer lender (P2P) and the first in the USA to register its offering as securities with SEC (Securities and Exchange Commission).  LC intersects technology and finance and with this approach there is the possibility of progressing a more stream-lined business model with an on-line offering and no physical facilities.  In addition they use machine learning to develop a risk management model that can eliminate lending bias and also focus on the most profitable customer base.

#### LC also has two distinct customer bases.  The loan customer and then the lenders.  They are a signifiant network of people providing the micro-loans and as a consequence of this model it is essential that business information is transparent enough for lenders to make informed decision about the percentage of defaults but also the rate of return for their funds. 

#### This data and the subsequent model is a good example of how DDD (Data Driven Decisions) can be used to ensure there is a more transparent approach to accessing personal finance and also investing in new ways.  

#### The data set is from Kaggle and is collected from 2007-2018 accepted applications.


#   Ethical consideration in the development of Machine Learning Models

#### When gathering personal data consumer data protection legislation needs to be considered, respected and the legislation followed.

#### Secondly, when considering the use of machine learning it is important to be transparent with customer and investors about the use of machine learning during the decision-making process. As noted in the 2021 annual reports from LC their marketing materials are clear that the company use over 150million cells of data to support the use of artificial intelligence-driven credit decisions and machine learning models are being used across the customer lifecycle to potentially expand access to credit.

#### Diversity and inclusion in the development of a model is also important as bias can be introduced if the machine learning models are limited by the data that it gets to train on.  Developing a team with a wide range of experience and from a diverse range of backgrounds is important to progress this.  I would go so far to say that it is particularly important to engage people with direct experience of some of the challenges of restricted access to credit due to personal or work situations.

#### I would recommend that further work be done on some of the new test models “What If” by Google so that the models can be challenged using a number of other data inputs.


#   Data Preparation and Cleaning
```{r}
view(lending_club_loans)
lending_club_loans <- clean_names(lending_club_loans)

```

```{r}
lendingverification <- lending_club_loans%>%
select(verification_status, loan_status)
view(lendingverification)
glimpse(lendingverification)
```

```{r}
lendingverification%>%
  count(loan_status, verification_status)

```

```{r}
ggplot(lendingverification)+
  geom_bar(aes(x = verification_status, fill = loan_status))+ 
  coord_flip()+
labs( x= "Verification Status",
        y= "Number of Loans",
        title = "Lending Club verification Status")
```
#### From the total number of charged off loans 5,653 we can see that 38% of the loans are not verified.  Increaseing the number of verification prior to approval 


```{r}
loans1 <- lending_club_loans %>%
  select(loan_amnt,term,int_rate,sub_grade,home_ownership,annual_inc,purpose, loan_status,emp_length, dti)
```


```{r}
ggplot(loans1)+
  geom_bar(aes(x = emp_length, fill = loan_status))+ 
  coord_flip()+
labs( x= "Length of Employment",
        y= "Number of Loans",
        title = "Lending Club Length of Employment")
```


#  View the structure of the data
```{r}
glimpse(loans1)
```
```{r}
str(loans1)
```


```{r}
summary(loans1)
```


```{r}
loans2 <- loans1 %>%
  filter(!is.na(loan_amnt))
loans2

loans3 <- loans2 %>%
  filter(!is.na(annual_inc))

```




```{r}
summary(as.factor(loans2$term))
```
```{r}
summary(as.factor(loans2$sub_grade))
```

```{r}
summary(as.factor(loans2$dti))
```

```{r}
summary(as.factor(loans2$purpose))
```


```{r}
summary(as.factor(loans2$loan_status))
```


```{r}
glimpse(loans3)
```
#Further cleaning of the data here at this point to drop the other loan status information.

```{r}
loansdropped <-loans3 %>%
  filter(loan_status %in% c("Charged Off","Fully Paid"
                    ))
loansdropped %>%
  count(loan_status)
```


```{r}
  ggplot(loansdropped) +
  aes(x = emp_length, fill = loan_status)+
  geom_bar(aes(x= loan_status), fill = "dark blue")+
labs( x= "Loan Status",
        y= "Number of Loans",
        title = "Lending Club Status of Loans Data")
```

```{r}
ggplot(loansdropped) +
  aes(x = term, fill = loan_status)+
  geom_bar(aes(x= loan_status), fill = "dark blue")+
labs( x= "Loan Status",
        y= "Number of Loans",
        title = "Lending Club Status of Loans Data")
```




```{r}
ggplot(loansdropped)+
  aes(x = emp_length, fill = loan_status)+
     geom_bar()+
    scale_fill_brewer(palette = "Dark2")+
  coord_flip()+
labs( x= "Length of Employment",
        y= "Number of Loans status",
        title = "Lending Club Length of Employment")
```

```{r}
ggplot(loansdropped)+
  aes(x = purpose, fill = loan_status)+
     geom_bar()+
    scale_fill_brewer(palette = "Dark2")+
  coord_flip()+
labs( x= "Purpose of the Loan",
        y= "Number of Loans  and Status of Loan",
        title = "Lending Club Purpose of Loan and Status")
```

```{r}
ggplot(loansdropped)+
  aes(x = sub_grade, fill = loan_status)+
     geom_bar()+
    scale_fill_brewer(palette = "Dark2")+
  coord_flip()+
labs( x= "Loan Grade",
        y= "Number of Loans and Status of Loan",
        title = "Lending Club Loan Grade and Status of Loan")
```

```{r}
ggplot(loansdropped)+
  aes(x = loan_amnt, fill = loan_status)+
     geom_bar()+
    scale_fill_brewer(palette = "Dark2")+
  coord_flip()+
labs( x= "Loan Grade",
        y= "Number of Loans and Status of Loan",
        title = "Lending Club Loan Grade and Status of Loan")
```

#  Key Take Aways

#### Total of 42,531 individual loan accounts with 5,653 charged off (meaning that the lender has written off the debt and the account is closed) and 33,586 fully paid off loans.

#### Majority of the loans are for 36 months (74%) and (26%) and for 60 months.

#### The LC internal credit rating ranges from A to G with 5 subcategories (a total of 35 credit ratings). A1 being the least likely to default and G5 the catergory with the highest risk profile.  From the data there is an uptick in the defaults at the D3 point onwards until we reach G5.

#### Debt consolidation (19,776 loans) 46% of loans followed by paying off credit card debt (5,477) 13% and then other (4425) 10% are the top three purposes of the loan request. It should be noted that this is self reported and caution should be exercised when considering the validity of this information.

#### While debt consolidation could arise from various other sources, such as auto or home equity lines, loans from these sources are secured and, hence, considerably different than unsecured credit like LC


# Analysis

#### Before building a model i think it is important to consider the key characteristics of a loan that has been defaulted on so that we are in a stronger position to understand what are the strong predictors of a loan defaulting.  I have undertaken further industry reading to inform this work.  A list of sources is at the end of the Markdown.


#### Her is will visual and explore the characteristics of the defaults loans only to uderstand more about the type of loan that is written off.



```{r}
default1 <-loansdropped %>%
  filter(loan_status == "Charged Off")
default1
```

```{r}
glimpse(default1)
```
```{r}
summary(default1)
```


```{r}
default1 %>%
  group_by(term, term_count = n())
```


```{r}
default1%>%
  count(loan_status)
```
```{r}
glimpse(default1)
```

```{r}
default1%>%
  select(term)%>%
count()
```



```{r}
defaulthome <-default1 %>% 
  group_by(home_ownership) %>%
  summarise(n = n ())
view(defaulthome)
summarise(defaulthome)
```


```{r}
default_arranged <- default1 %>%
  select(loan_amnt, term)%>%
  arrange(desc(loan_amnt))
```

```{r}
summary(default_arranged)
```


#### Data set for default loans is 5,653 and on average the default amount is $12,134 - across the business this could be an estimated value of $68,593,502 so any reduction in the number of defaulting loans by say 1% could be $68,000 additional revenue.




```{r}
ggplot(default1) +
  geom_bar(aes(x= term), fill = "light blue")+
labs( x= "Term of Loan",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data")
```


#### Loans with a longer term of 60 months make up a higher percentage of the defaults loans - we could conclude that the longer the term the higher the exposure to changes in a persons financial status e.g employment or other personal circumstances. Historically, only the most financially secure individuals would be eligiable for a longer term loan.

#### For the purpose of this work the loan term could be considered a factor in the probability of a customer defaulting.


```{r}
ggplot(default1) +
  geom_bar(aes(x= loan_amnt), fill = "dark blue")+
  coord_flip()+
labs( x= "Loan Amount in $",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data")
```
#### Here you can see four spikes in the data where there are more defaults on certain values of loans - $7,000, $18,000, $25,000 and then the max of $35,000 - the loan size could be considered a factor in the probabilty of a customer defaulting and risk profile increased (e.g interest rates) increased at these particular values.



```{r}
ggplot(default1) +
  geom_bar(aes(x=home_ownership), fill = "dark blue")+
   labs( x= "Houseing Status",
        y= "Number of Default Loans",
        title = "Lending Club Loans Data",
        subtitle = "Houseing Status")
```
#### An applicants housing status can also be considerd a factor in default loans.  There is a a higher percentage of renters - unsurprising and also might be considered a higher risk when lending as a credit check cannot be secured from other traditional lender. Furthermore those with a mortage may be in a position to get a loan that is secured against their property. Housing status could be considered a factor in the probability of a customer defaulting.

#### There is a slight increase in the number of people defaulting who are renters with 50% renting and 41% having a mortage.  For the purpose of the model home status could be consider a factor in the probabilty of a customer defaulting.

```{r}
default_dti <-default1 %>%
  select(dti, term)%>%
  arrange(desc(dti))

```


```{r}
ggplot(default_dti) +
  geom_bar(aes (x = dti, fill = term), stat = "count")+
  labs( x= "Count %",
        y= "Debt to Income Ratio",
        title = "Lending Club Loans Data v Term",
        subtitle = "DTI %")
 
```



```{r}
ggplot(default1) +
  geom_bar(aes(x=dti), fill = "light blue")+
   labs( x= "Debt to Income Ratio %",
        y= "Number of Default Loans",
        title = "Lending Club Loans Data",
        subtitle = "DTI %")
```



```{r}

```



```{r}
ggplot(default1) +
  geom_bar(aes(x= term, fill = purpose))+
  labs( x= "Term of Default Loan",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data",
        subtitle = "Term & Purpose",
        fill = "Purpose of Loan")
```
#### Debt consolidation, payment of credit card debts and "other" continue to be the top three reasons for a loan and for a customer then defaulting.  The purpose of a loan could be considered a factor in the chance of a customer defaulting and debt consolidation being a higher risk factor than for education or home improvements.

#### Limiting exposure to the loans being used for this purpose may be a consideration in the future or offering a higher interest rate to mitiage the risk.




```{r}
ggplot(default1) +
  geom_bar(aes(x= int_rate, fill = term))+
  labs( x= "Interest Rate",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data",
        subtitle = "Term & Purpose",
        fill = "Term")
```


  

```{r}
ggplot(default1) +
  geom_bar(aes(x= term, fill = annual_inc))+
  labs( x= "Term of Default Loan",
        y= "Number of Default Loans",
        title = "Lending Club Default Loans Data",
        subtitle = "Term & Purpose",
        fill = "Purpose of Loan")
```



#loan status needs to be turned into a numerical column 0 or 1 for purpose of building a predictive model.


# Prediction of Loan Defaults

#### My ultimate goal is the prediction of loan defaults from a given set of observations by selecting explanatory (independent) variables (also called feature in machine learning) that result in an acceptable model performance as quantified by a pre-defined measure. This goal will also impact our exploratory data analysis. 



```{r}

loansmodelling <-loansdropped %>%
  mutate(loan_status = recode(loan_status,"Fully Paid" = FALSE, "Charged Off" = TRUE)) 
loansmodelling
```


```{r}
ggplot(loansmodelling,
       aes(x=dti, y = as.integer (loan_status))) +
  geom_point(fill = "light blue")+
  geom_smooth()+
  coord_cartesian(ylim = c(0.05, 0.2))+
   labs( x= "Debt to Income Ratio %",
        y= "Number of Default Loans",
        title = "Lending Club Loans Data",
        subtitle = "DTI %")
```


```{r}
glimpse(loansmodelling)
```



```{r}
head(loansmodelling, 15)
```


#   Machine Model ~ 1 Logistic Regression

#### Predictive analysis using a machine learning model is an effective way of considering the relationships between a variable that may impact on a customer defaulting.

#### Now let's have a look at the relationships with loan status. I have split the variables into three sets, as the output of `ggpairs()` was slow and cramped when I include all variables at once. Note that a disadvantage of this approach is that it will be difficult to visually assess relationships between some predictor pairs if they fall into different split sets.

#### My ultimate goal is the prediction of loan defaults from a given set of observations by selecting explanatory (independent) variables (also called features in machine learning) that result in an acceptable model performance. This goal will also impact our exploratory data analysis. 

#### the data has been split into two sets identifying loan amount, term and home ownership against the loan status.  The final set is annula income, the purpose of the loan and dti (debt to income ratio)

#### GGpairs is then used to see the predictive relationships with the variables.



```{r}
loans_split1 <- loansmodelling %>%
  select(loan_status, loan_amnt,term,home_ownership)
```


```{r}
loans_split2 <- loansmodelling %>%
  select(loan_status, annual_inc, purpose, dti)
```

#Now develop the connection with ggpairs

```{r message=FALSE}
library(GGally)
ggpairs( loans_split1)
```
#

```{r}
score_plot1 <- ggplot(loans_split1) +
  geom_jitter(aes(y = loan_amnt, x = as.integer(loan_status)), shape = 1, 
              position = position_jitter(width = 0.05)) + 
  xlab("Loan Status") + scale_x_continuous(breaks=seq(0, 1,1))

score_plot1
```


```{r}
score_plot2 <- ggplot(loans_split1) +
  geom_jitter(aes(y = term, x = as.integer(loan_status)), shape = 1, 
              position = position_jitter(width = 0.05)) + 
  xlab("Loan Status") + scale_x_continuous(breaks=seq(0, 1,1))

score_plot2
```


```{r message=FALSE}
library(GGally)
ggpairs(loans_split2)
```


```{r}
score_plot3 <- ggplot(loans_split2) +
  geom_jitter(aes(y = dti, x = as.integer(loan_status)), shape = 1, 
              position = position_jitter(width = 0.05)) + 
  xlab("Loan Status") + scale_x_continuous(breaks=seq(0, 1,1))

score_plot3
```


```{r}
loan_data_logreg_model <- glm(loan_status ~ dti+ loan_amnt+ term +log(annual_inc), data = loansmodelling, family = binomial(link = 'logit'))
loan_data_logreg_model
```
#### dti, loan amount and term are all good predictors of the higher chance of defaulting but annual income returns a negative and is not that strong a predictor.

#### But we need to think carefully about the statistical significance of each predictor. A good first step in this direction is to examine the p-value of each predictor. Probability value

#### A p-value is a statistical measurement used to validate a hypothesis against observed data. I measures the probability of obtaining the observed results, assuming that the null hypothesis is true. The lower the p-value, the greater the statistical significance of the observed difference


```{r}
tidy_out <- clean_names(tidy(loan_data_logreg_model))
tidy_out
```

#### From this we can see that dti and term are strong predictors with loan amount and annual income not so much. 


```{r}
loan_data_logreg_model2 <- glm(loan_status ~ purpose + home_ownership + term +log(annual_inc), data = loansmodelling, family = binomial(link = 'logit'))
loan_data_logreg_model2
```


```{r}
tidy_out <- clean_names(tidy(loan_data_logreg_model2))
tidy_out
```


#### From this we can see that when looking at the purpose of a loan debt consolidation, "other" and a "major purchase"  are strong predictors with medical bills and moving being less of a factor.



#   Recommendations


#### The following are some suggestions to consider in relation to how LC might alter their existing approach to lending so that defaults are reduced.



#### Reduce exposure to longer term loans.

#### Reduce exposure to loans graded D1 and beyond.

#### Model three-tier lending term as opposed to a two-tier.

#### Reduce exposure to debt consolidation/credit card loans.

#### Reduce exposure to customers with a DTI of between 22 - 25% and above.

#### Increase verification of loans information prior to approval.

#### Review the gradiant at which the Loan approval, based on risk profile and mitigation of the risk, and further increase the interest rate.




#   sources

#### https://www.lendingclub.com/loans/resource-center/calculating-debt-to-income

#### https://www.lendingclub.com/foliofn/rateDetail.action

#### https://www.nclc.org/resources/predatory-installment-lending-in-the-states-2021/

https://www.federalreserve.gov/releases/chargeoff/delallsa.htm   - predominately to look at deliquency rates (charge off) in P2P but only commercial banks available.

#### https://www.federalreserve.gov/econres/notes/feds-notes/do-marketplace-lending-platforms-offer-lower-rates-to-consumers-20181022.html - considering if P2P offer more competative rates to people.

#### https://moneyclubber.com/blog/how-high-is-the-default-rate-on-peer-to-peer-lending/

#### https://www.fca.org.uk/







